---
title: "NY Shooting Data"
author: "Antony Sequeira"
date: "2/9/2022"
output: pdf_document
---

# Overview
For fulfillment of DTSA-5301 finals *NYPD Shooting Incident Data Report* part of the assignment.

The data is free government data about shooting incidents in New York city. 
The goal is to get, cleanup, analyse, and present the NYPD shooting data.
The main goal of the analysis is to find out how the incidents relate to the different factors.

## Following script installs the required libraries in Mac OS
This section can be copied to a file or input into R console.  
You could also download it from my repo at https://github.com/asequeir-edu-2022/dtsa5301final

```
#!/usr/bin/env Rscript
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

print("Installing R libraries")
install.packages("chron")
install.packages("tidyverse")
install.packages("tinytex")

tinytex::install_tinytex()
```

## Load libraries
```{r setup, echo=TRUE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, results = "hide", warning = FALSE)

library(tidyverse, warn.conflicts=F, quietly=T)
library(lubridate, warn.conflicts=F, quietly=T)
library(chron, warn.conflicts=F, quietly=T)
```

```{r reset, echo=TRUE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, results = "echo")
```

# Data source
Data is downloaded from CSV link in  
https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic
```{r load_data, echo=TRUE, results=FALSE, message=FALSE}
ny_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
ny_data_raw <- read_csv(ny_url)
```


## Clean up data
We clean up the data mainly through the following three operations:  
- variables to factor for appropriate columns  
- date types from strings  
- remove unneeded columns  

```{r clean_data, echo=TRUE}
ny_data <- ny_data_raw %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  mutate(OCCUR_TIME = chron(times=OCCUR_TIME)) %>%
  mutate(BORO = factor(BORO)) %>%
  mutate(PRECINCT = factor(PRECINCT)) %>%
  mutate(PERP_AGE_GROUP = factor(PERP_AGE_GROUP)) %>%
  mutate(PERP_SEX = factor(PERP_SEX)) %>%
  mutate(PERP_RACE = factor(PERP_RACE)) %>%
  mutate(VIC_AGE_GROUP = factor(VIC_AGE_GROUP)) %>%
  mutate(VIC_SEX = factor(VIC_SEX)) %>%
  mutate(VIC_RACE = factor(VIC_RACE)) %>%
  select (-c(JURISDICTION_CODE, LOCATION_DESC, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, INCIDENT_KEY))
```

## Missing data columns and plans to handle them
There is missing data in the following columns:
```{r missing_data, echo=TRUE}
names(which(colSums(is.na(ny_data_raw)) > 0))
```
Missing data in factor columns `PERP_SEX` etc. are handled already as a factor.  
I do not plan to use `JURISDICTION_CODE` and `LOCATION_DESC`.  
So, for this data, nothing more needs to be done for missing data handling. 

### Summary of the cleaned up data

```{r summary, echo=TRUE}
summary(ny_data)
```
Turning columns into factors shows total counts for columns such as `BORO`.  
The summary also shows the breakdown of total incidents by factor type columns such as `PERP_AGE_GROUP`.  


## Visualization and Analysis

### Generate necessary analysis ready data.  
Prepare a few different slices of the data for visualizations.  

```{r try1}
ny_data_sum <- ny_data %>%
  group_by(BORO, VIC_AGE_GROUP) %>% 
  mutate(BORO_BY_VIC_AGE = n()) %>%  # occurrences by victim age group by boro
  ungroup()  %>%
  group_by(month = lubridate::floor_date(OCCUR_DATE, "month")) %>%
  mutate(month_sum = sum(n())) %>% # occurrences by month
  ungroup() %>%
  group_by(month, BORO) %>%
  mutate(month_by_boro = sum(n())) %>%
  ungroup()
```

```{r sum2}
ny_data_sum2 <- ny_data %>%
  group_by(BORO, VIC_AGE_GROUP) %>% 
  summarize(BORO_BY_VAG = n(), .groups = 'drop') %>%
  ungroup()
```

```{r sum3}
ny_data_sum3 <- ny_data %>%
  group_by(month = lubridate::floor_date(OCCUR_DATE, "month")) %>%
  summarise(month_sum = sum(n())) %>% # occurrences by month
  ungroup()
```

## Visualizations
### Variations by the month of the year
We plot the total incidents by month.  
```{r plot1}
ny_data_sum3 %>%
ggplot(aes(x = month, y = month_sum)) +
geom_line(aes(color = "month")) +
geom_point(aes(color = "month")) +
geom_line(aes(y=month_sum, color = "month_sum")) +
geom_point(aes(y=month_sum, color = "month_sum")) +
xlab("Year/Month") +  ylab("Incidents") +
labs(title = "Incidents in NY by month")
```
The above plot shows that the number of incidents vary seasonally.  
To see the 2020 peak more clearly, we plot a shorter time span.  
```{r plot2}
ny_data_sum3 %>%
  filter(year(month) > 2019) %>% 
  ggplot(aes(x = month, y = month_sum)) +
  geom_line(aes(color = "month")) +
  geom_point(aes(color = "month")) +
  geom_line(aes(y=month_sum, color = "month_sum")) +
  geom_point(aes(y=month_sum, color = "month_sum")) +
  xlab("Year/Month") +  ylab("Incidents") +
  labs(title = "Incidents in NY by month for years 2019 onwards")
```

This shows a clear peak in July of 2020.  

### Variations by borough
Generate the counts by borough.  
```{r plot4}
ny_data_sum %>%
  group_by(month_by_boro) %>%
  ggplot(aes(x=month, y=month_by_boro, group=BORO, color=BORO)) +
  geom_line() + 
  xlab("Year/Month") +  ylab("Incidents") +
  labs(title = "Incidents in NY by month by Boro")
```

Get a smaller time span to see a zoomed in view.  
```{r plot5}
ny_data_sum %>%
  filter(year(month) > 2015) %>% 
  filter(year(month) < 2019) %>%
  group_by(month_by_boro) %>%
  ggplot(aes(x=month, y=month_by_boro, group=BORO, color=BORO)) +
  geom_line() + 
  xlab("Year/Month") +  ylab("Incidents") +
  labs(title = "Plot fewer years to show peaks")
```

## Analysis
The plots show the following:  

- seasonal peaks mostly in summer
- higher levels of incidents based on the boro - Staten Island is lowest and Bronx and Brooklyn seem to be the higher end.
- the incidents show unusual higher numbers in first quarter of 2020


## Questions raised by the visualization and analysis (to be investigated)
- population of boros (Staten Island might have much smaller population) may be too different
- factors not in data such as income
- number of police officers per person

## Bias
There could be multiple sources of bias in the NY shooting data

- the data collection may be biased, it is possible that not all shootings are reported
- the standard race categories may not reflect the reality of the NY demographics

I have tried to focus on the boro and seasonality of the data to reduce bias.

## Modeling
```{r}
ny_data_doy <- ny_data_sum %>%
  filter(year(OCCUR_DATE)< 2020) %>% # avoiding covid years
  group_by(doy = yday(OCCUR_DATE)) %>%
  mutate(doy_sum = sum(n())) %>%
  ungroup()

mod <- lm(doy_sum ~ doy, data=ny_data_doy)
summary(mod)

ny_data_pred <-  ny_data_doy %>% mutate(doy_sum_pred = predict(mod))

ny_data_pred %>% ggplot() +
  geom_point(aes( x = doy, y = doy_sum), color = "blue") +
  # compare with predicted
  geom_point(aes( x = doy, y = doy_sum_pred), color = "red") +
   ylab("daily incidents") + xlab("day of the year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) 

```


The plot clearly shows that the linear model does not fit the data of incidents given a day of the year.  
It looks like we need a different model (other than linear) to predict incidents given any day of the year. 

I hope to learn more about statistical modelling in future data science courses so I can model such data better.

## Session info
```{r, echo=FALSE}
sessionInfo()
```
