---
title: "COVID19 Data from John Hopkins"
author: "Antony Sequeira"
date: "2/9/2022"
output: pdf_document
---

# Overview
For fulfillment of DTSA-5301 finals *COVID19 dataset from the Johns Hopkins github site* part of the assignment.

## Following script installs the required libraries in Mac OS
This section can be copied to a file or input into R console.  
You could also download it from my repo at https://github.com/asequeir-edu-2022/dtsa5301final

```
#!/usr/bin/env Rscript
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

print("Installing R libraries")
install.packages("chron")
install.packages("tidyverse")
install.packages("tinytex")

tinytex::install_tinytex()
```

## Load libraries
```{r setup, echo=TRUE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, results = "hide", warning = FALSE)

library(tidyverse, warn.conflicts=F, quietly=T)
library(lubridate, warn.conflicts=F, quietly=T)
library(chron, warn.conflicts=F, quietly=T)
```

```{r reset, echo=TRUE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, results = "echo")
```


# Data source
The main data source is the github repository at
https://github.com/CSSEGISandData/COVID-19

We will use the data from the folder at  
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series

Two of the time series files are for the US confirmed cases and deaths, reported at the county level.  
They are `time_series_covid19_confirmed_US.csv` and `time_series_covid19_deaths_US.csv` respectively.

The report will focus on analyzing the data for United States only.

```{r load_data, echo=TRUE, results=FALSE, message=FALSE}
covid19_confirmed_url = "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
covid19_deaths_url = "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

covid19_confirmed_raw <- read_csv(covid19_confirmed_url)
covid19_deaths_raw <- read_csv(covid19_deaths_url)
```

## Data description and cleaning
The `Province_State` column contains the name of the US State and the `Admin2`
contains the County name for which daily data is collected.  
These are the main location identifiers in this data.  
The date columns are in long form and contain the cases and death counts.  

We clean up the data mainly through the following three operations:  
 - variables to factor for appropriate columns  
 - date types from strings  
 - remove unneeded columns  
 - pivot from long date columns

### Pivot long 

This converts the long form date columns to a single date column with values. 
```{r pivot_data, echo=TRUE}
covid19_confirmed <- covid19_confirmed_raw %>%
  pivot_longer(cols = -c(UID:Combined_Key),
                       names_to = "date",
                       values_to = "cases") %>%
  select(-c(Lat, Long_)) ## remove unwanted columns

covid19_deaths <- covid19_deaths_raw %>%
  pivot_longer(cols = -c(UID:Population),
                       names_to = "date",
                       values_to = "deaths") %>%
  select(-c(Lat, Long_)) ## remove unwanted columns

```

### Join cases and deaths data and change column types (factor, date)

We change date strings to dates and identifier kind of string columns to factors.  
```{r join_factor_data, echo=TRUE}
covid19 <- covid19_confirmed %>%
  full_join(covid19_deaths) %>%
  mutate(iso3 = factor(iso3)) %>%
  mutate(Admin2 = factor(Admin2)) %>%
  mutate(Province_State = factor(Province_State)) %>%
  mutate(Country_Region = factor(Country_Region)) %>%
  mutate(date = mdy(date)) %>%
  select (-c(UID, iso2, code3, FIPS))
```


## Missing data columns and plans to handle them
There is missing data in the following columns:
```{r missing_data, echo=TRUE}
names(which(colSums(is.na(covid19)) > 0))
```

I will ignore such data for this report by filtering.

```{r cruise_ships_remove, echo=TRUE}
covid19 <- covid19 %>%
  filter(iso3 == "USA") %>%  # filter non states for simplicity
  filter(!is.na(Admin2))     # ignore cruise ships
```

### Summary of the cleaned up data

```{r summary, echo=TRUE}
summary(covid19)
```

### filter data
I will remove data with zero cases for simplicity.  
```{r filter_data, echo=TRUE}
covid19 <- covid19 %>%
  filter(cases > 0)
```

### Data issues
There are deaths with population zero. This is because the deaths are recorded with unassigned county
(Alaska for example).
These will have to be removed if we need any division by population operation for any individual date.
I do not plan to do that.

## Visualization and Analysis

### Generate necessary analysis ready data.  

Generate per state aggregated data using grouping by location and summing the columns needed.

```{r}
covid19_by_state <- covid19 %>%
  group_by(Province_State, Country_Region, date) %>%
  summarise(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_million = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, 
         cases, deaths, deaths_per_million, Population) %>%
  ungroup()
```

### Plot 
This plot shows the accumulated cases and deaths over the whole of US.
The plot looks thick since each state gets a point per day.
```{r}
covid19_by_state %>%
  ggplot(aes(x = date, y = cases)) +
  
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in US", y = NULL)
```

Plotting for the state of California only.
```{r}
covid19_by_state %>%
  filter(Province_State == "California") %>%
  ggplot(aes(x = date, y = cases)) +
  
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in California", y = NULL)
```


Plot for the month of March, 2020 only. 
This shows the range of values on a given day across all the US states.
```{r}
covid19_by_state %>%
  filter (year(date) == 2020) %>%
  filter (month(date) == 3) %>%
  # filter (month(date) == 4 | month(date) == 5) %>%
  # filter (day(date) < 25) %>%
  ggplot(aes(x = date, y = cases)) +
  
  geom_point(aes(color = "cases")) +
  
  geom_point(aes(y = deaths, color = "deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in US in March of 2020", y = NULL)
```

## Summarize by monthly numbers

Aggregating the data by getting one point per month per state.

```{r}
covid19_by_state_month <- covid19_by_state %>%
  group_by(Province_State,  month = lubridate::floor_date(date, "month")) %>%
  summarise(cases = min(cases), # use value from the 1st of the month
            deaths = min(deaths), 
            Population=max(Population),
            cases_per_million = cases * 1000000 / Population,
            deaths_per_million = deaths * 1000000 / Population
            ) %>%  # occurrences by month
  ungroup()
```


Plotting this data shows the variations in the pandemic across the US states.

```{r}
covid19_by_state_month %>%
  group_by(month) %>%
  ggplot(aes(x=month, y=cases, group=Province_State, color=Province_State)) +
  geom_line() + 
  labs(title = "Cases by US State", y = NULL)
```

### Analyze a single day

Get data for a single day and show the top and bottom states by cases.


```{r}
all_data = covid19_by_state_month %>% 
                   filter(month == as.Date("2022-01-01")) %>%
                   select (-c(month, cases_per_million))

```

```{r}
head(all_data %>% arrange(cases))
```

```{r}
tail(all_data %>% arrange(cases))
```

Same data ordered by `cases_per_million` shows a different ordering.

```{r}
all_data = covid19_by_state_month %>% 
                   filter(month == as.Date("2022-01-01")) %>%
                   select (-c(month, deaths_per_million))

```

```{r}
head(all_data %>% arrange(cases_per_million))
```
```{r}
tail(all_data %>% arrange(cases_per_million))
```

## Data for model fitting
Generate daily cases from the cumulative.
```{r}
covid19_by_state <- covid19_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths)
         )
```

Plot them for a day to see they look right.
```{r}
covid19_by_state %>%
  filter (year(date) == 2020) %>%
  filter (month(date) == 3) %>%

  ggplot(aes(x = date, y = new_cases)) +

  geom_point(aes(color = "new_cases")) +
  
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in US new cases and new deaths", y = NULL)
```

Filter data for California only for a better fit.

```{r}
california_data = covid19_by_state_month %>% 
  filter(cases > 0) %>%
  filter(Province_State == "California") 
```

Model the data using linear model to fit deaths to cases.
```{r}
model <- lm(deaths ~ cases, data = california_data)
summary(model)
california_data_pred  <- california_data %>%
  mutate(pred = predict(model))
```

Plot the model data with predicted values.

```{r}
california_data_pred %>%
  ggplot(aes(x = cases, y = deaths)) +
  
  geom_point(aes(color = "deaths")) +
  
  geom_point(aes(y = pred, color = "pred")) +
  
  # scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in US model fit - predict deaths from cases in California", y = NULL)
```

## Bias
There could be multiple sources of bias in the COVID19 data:

- The data collection is likely biased since this data is collected over many jurisdictions with possibly different rules of reporting.
- The data also spans a time window that 2 years where data collection process may not be uniform.
- I have picked California to fit a model. It is possible that a linear model may not fit for all states.

## Session info
```{r, echo=FALSE}
sessionInfo()
```



